---
title: "ukrainians"
format: html
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# set code chunk option defaults
knitr::opts_chunk$set(
  # display code as types
  tidy = FALSE, 
  # slightly smaller code font
  size = "small",
  # do not display messages in PDF
  message = FALSE,
  # set default figure width and height
  fig.width = 5, fig.height = 3) 
```

```{r, include=FALSE, warning=FALSE}

# Load necessary packages
if (!require('pacman')) install.packages('pacman'); library('pacman')
pacman::p_load(viridis, tidyverse, ggplot2, lubridate, knitr, dplyr, ggpubr, 
              janitor, tidyr, kableExtra, ggcorrplot, ggthemes, zoo, forcats,
              gridExtra, grid, Hmisc, FSA, dunn.test, readxl, RColorBrewer,
              patchwork, cowplot, pROC, rms, vcd, pscl)
# Load data
data_ee <- readRDS("Data/data_ee")
combined_df <- readRDS("Data/combined_df")
combined_data <- readRDS("Data/combined_data")
```

*Introduction*

The inflow of more than 8 million Ukrainian refugees into Europe has introduced both challenges and opportunities for labour market outcomes. People fleeing the war in Ukraine are allowed to live and work in the European Union for up tp 3 years under the Temporary Protection Status. Although the education and qualification levels of Ukrainian refugees are generally high, difficulties such as lack of language skills, childcare services and credential recognition processes pose significant barriers to employment. Instances of ethnic discrimination in the labor market have been reported, highlighting additional challenges faced by Ukrainian refugees (Londar et al., 2024). According to Preut, expanding language courses, childcare services, and accelerating credential recognition can enhance employment likelihood (Preut, 2023).

*Data and methods*

This study employs Multi-Sectoral Needs Assessments (MSNA) of 2023, which provide data on Ukrainian refugees' needs and priorities in Estonia, Slovakia, Poland, Romania, Moldova, and Czechia. The assessment employs both household-level and individual-level data collected through structured surveys in all the countries.

**Data Wrangling and Preparation**

To enable cross-regional comparisons, variables that were consistently present across all six national datasets were selected and merged into a single dataset. Questions allowing multiple responses (e.g., types of support received, reasons for unemployment, barriers to accessing services) were transformed using one-hot encoding into a series of binary indicator variables (1 = selected, 0 = not selected) to show each of the possible response categories in disaggregated form. Where categorical responses varied slightly across countries (e.g., differences in wording or coding), the values were recoded into a uniform format for comparison. 

For the convenience of exploratory data analysis (EDA) and model building, the outcome variable (employment status) was defined as follows:

- Respondents were classified as employed (1) if they reported any form of employment.
- All others (unemployed, in education, retired, or not working due to other factors) were classified as unemployed (0).

This two-category outcome variable, Employed, was used as the dependent variable for all the following regression and classification models.

Based on the proportions of missing data and their impact on model performance, non-available observations were excluded or imputed. Individual aggregated dataframes were created so that the distribution of binary-coded multi-response variables could be visualized and country-level comparisons could be made.

**Statistical Modeling**

Logistic regression and random forest classification were applied in an attempt to identify influential factors concerning employment among Ukrainian refugees:

- Random Forest Classifier

A random forest model was trained to account for potential non-linearities and interactions between variables. The model was trained using predictors selected based on feature importance scores. Its performance was evaluated using classification measures such as accuracy, precision, recall, and area under the ROC curve (AUC). Random forest allowed for selection of the most significant predictors of job-finding based on mean decrease in impurity (Gini importance) and permutation importance.

- Logistic Regression

Binary logistic regression analysis was used to estimate the probability of employment (Employed = 1). Independent variables included demographic and socio-economic ones such as age, education level, country of residence, language skills, accommodation, and social services consumption of social services, among others. Odds ratios were estimated to define direction and strength of associations.

*Descriptive Statistics*

**Demographics**

The bar chart displays the distribution of survey respondents by age group (18–34, 35–59, 60+) across six Central and Eastern European countries. Across all countries, the 35–59 age group consistently has the highest number of respondents, followed by the 18–34 group, with the 60+ group being the least represented. Romania stands out with a particularly high concentration of respondents in the 35–59 group compared to the other age categories. Estonia shows a more balanced distribution between the younger and middle-aged groups, though the same pattern persists.


```{r}
# Custom plotting function
plot_employment_vs_vars <- function(data, vars, country_name) {
  # Filter for country
  df <- data |> 
    filter(country == country_name) |> 
    select(employed_binary, all_of(vars)) |> 
    drop_na()

  df$employed_binary <- factor(df$employed_binary, labels = c("Unemployed", "Employed"))

  for (var in vars) {
    # Convert variable to factor if it's not numeric
    if (!is.numeric(df[[var]])) {
      df[[var]] <- as.factor(df[[var]])
    }

    # Generate plot
    p <- ggplot(df, aes(x = employed_binary, fill = as.factor(.data[[var]]))) +
      geom_bar(position = "fill") +
      scale_y_continuous(labels = scales::percent_format()) +
      labs(
        title = paste("Employment vs", var, "in", country_name),
        x = "Employment Status",
        y = "Proportion",
        fill = var
      ) +
      theme_minimal()

    print(p)  # You can use ggsave() here to save plots instead of printing
  }
}
```

```{r, fig.width=6, fig.height=4, warning=FALSE, echo=FALSE}
# Function to create a clean proportion table
make_prop_table <- function(variable, na.rm = FALSE) {
  if (na.rm) {
    variable <- variable[!is.na(variable)]
  }

  tab <- table(variable)
  
  tibble::tibble(
    Category = names(tab),
    Count = as.numeric(tab),
    Percent = round(100 * prop.table(tab), 2)
  )
}

make_prop_table(data_ee$introduction_resp_age)

gg <- theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 7)
  )

combined_data |>
  filter(!is.na(introduction_resp_age), !is.na(country)) |>
  mutate(introduction_resp_age = factor(introduction_resp_age,
                                        levels = c("18-34", "35-59", "60+"))) |>
  count(country, introduction_resp_age) |>
  group_by(country) |>
  mutate(Percent = n / sum(n) * 100) |>
  ggplot(aes(x = country, y = Percent, fill = introduction_resp_age)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    x = "Country",
    y = "Percent of Respondents",
    fill = "Age Group",
    title = "Age Distribution of Respondents by Country"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text.y = element_text(size = 10),
    plot.title = element_text(hjust = 0.5)
  ) +
  guides(fill = guide_legend(title.position = "top", title.hjust = 0.5)) +
  scale_fill_brewer(palette = "Dark2")

```

While the data on education of Ukrainian refugees is limited, the snapshot of age distribution reveals that the majority of arrivals are part of labor force. Notably, in Estonia, 49.46% of arrivals hold a vocational and specialization education, followed by around 38% of refugees with Master or Bachelor degrees. People with secondary education are moderately represented, but there are very few people with PhD education. The largest group is employed, indicating that many refugees were active in the workforce before displacement. Other categories include students, self-employed, and those in housekeeping.

The educational categories range from primary and secondary education to higher levels such as Bachelor's, Master's, and PhD degrees, as well as technical or vocational training and no formal education. "Technical/Vocational" category appears frequently across multiple sections, suggesting a significant portion of the workforce in these regions may have either technical training. This trend could indicate a reliance on skilled labor or informal employment in certain areas. The number of people with Bachelor's degrees is almost the same across all countries and ranges from 10 to 20%. Slovakia and Romania have the highest proportion of people with Master's degrees. There are very few people with no education in all the countries. Slovakia and Czechia appear the most prominent for respondents with PdD degrees. 

```{r, fig.width=7, fig.height=5, warning=FALSE, echo==FALSE}

combined_data |>
  filter(!country %in% c("Hungary", "Bulgaria", "construct", NA)) |>
  mutate(demographics_educ_level = case_when(
    grepl("tech", demographics_educ_level, ignore.case = TRUE) ~ "Technical/Vocational",
    grepl("sec", demographics_educ_level, ignore.case = TRUE) ~ "Secondary",
    grepl("pri", demographics_educ_level, ignore.case = TRUE) ~ "Primary",
    grepl("phd", demographics_educ_level, ignore.case = TRUE) ~ "PhD",
    grepl("master", demographics_educ_level, ignore.case = TRUE) ~ "Master's",
    grepl("bachelor", demographics_educ_level, ignore.case = TRUE) ~ "Bachelor's",
    grepl("no_edu", demographics_educ_level, ignore.case = TRUE) ~ "No Education",
    grepl("prefer", demographics_educ_level, ignore.case = TRUE) ~ "No Answer",
    grepl("spec", demographics_educ_level, ignore.case = TRUE) ~ "Specialization",
    TRUE ~ demographics_educ_level
  )) |>
  filter(!is.na(demographics_educ_level) & demographics_educ_level != "No Answer") |>
  count(country, demographics_educ_level) |>
  group_by(country) |>
  mutate(prop = n / sum(n) * 100) |>
  ggplot(aes(x = fct_infreq(demographics_educ_level), y = prop, fill = demographics_educ_level)) +
  geom_col() +
  facet_wrap(~country) +
  xlab("Education Level") +
  ylab("Percent") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = guide_legend(title = "Education Level")) +
  scale_fill_brewer(palette = "Dark2")

```



**Economic capacity and Aid**

The faceted bar chart indicates the sources of declared income. The most apparent trend is the dominant role of "employment in the host country" as the principal source of income in most countries, especially in Czechia, Estonia, and Slovakia, where it accounts for over 40–50%. Romania has a more balanced split, with employment percentages, other sources, and maintained employment from Ukraine all being similar, however, it also has the highest number of people with no income. Remote work and remittances in Romania and Poland respectively also arise more than anywhere else. Across all countries, rates like "no answer", and "no income" are low. 

```{r, fig.height = 7, fig.width=8, echo=FALSE}

#Aggregated data

combined_df |>
    filter(Source == "income") |>
    mutate(Percent = round(Total_Proportion * 100, 2))|>
    select(Component, Percent, Country) |>
    group_by(Country) |>
    arrange(desc(Percent), .by_group = TRUE) |>
    slice_head(n = 5) |>
    ungroup() |> 
  ggplot(aes(x = reorder(Component, -Percent), 
             y = Percent, fill = Component)) +
  geom_col(show.legend = TRUE) +
  facet_wrap(~Country, scales = "free_x") +
  xlab("Income Source") + ylab("Percent") +
  ggtitle("Comparison of Reported Income Sources by Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = guide_legend(title = "Income Source"))+
  scale_fill_brewer(palette = "Dark2")

```


The chart shows that the most commonly reported reason for not experiencing employment problems is "None," particularly in Moldova and Slovakia, where over 40% of respondents indicated no difficulties. In contrast, Estonia stands out with a high incidence of the "Language barrier" (around 37%), suggesting it is a major obstacle there. Across all countries, language barriers and the fact that many are “Not looking for work” are consistently among the top reasons cited, though to varying degrees. "Lack of skills" and "Lack of recognition" are less frequently mentioned, typically below 10%. Moldova, Romania, and Poland also report relatively high percentages for "Not looking for work" and "Insufficient opportunities," indicating structural and motivational challenges.

```{r, fig.height = 7, fig.width=8, echo=FALSE}

#Aggregated data

combined_df |>
  filter(!Country %in% c("Hungary", "Bulgaria", "construct", NA)) |>
    filter(Source == "job") |>
    mutate(Percent = round(Total_Proportion * 100, 2)) |>
    select(Component, Percent, Country) |>
    group_by(Country) |>
    arrange(desc(Percent), .by_group = TRUE) |>
    slice_head(n = 5) |>
    ungroup() |> 
  ggplot(aes(x = reorder(Component, -Percent), 
             y = Percent, fill = Component)) +
  geom_col(show.legend = TRUE) +
  facet_wrap(~Country, scales = "free_x") +
  xlab("Income Source") + ylab("Percent") +
  ggtitle("Problems with Finding Employment") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = guide_legend(title = "Reasons")) +
  scale_fill_brewer(palette = "Dark2")
```

The graph compares employment activity of Ukrainian refugees prior to displacement with their current activities in host countries. In Ukraine, the vast majority of refugees across all countries were employed, with employment rates ranging from approximately 60% to over 70%. Housekeeping and retirement were the next most common categories, but far less prevalent. In contrast, in host countries, employment drops significantly, while unemployment, studying, and housekeeping rise sharply. For example, in Poland, Romania, and Slovakia, a substantial portion - around one-third - of refugees are now unemployed. Studying categories also grow considerably, particularly in Romania and Moldova. This shift indicates a major transition for many refugees, from stable employment in Ukraine to navigating unemployment or reskilling opportunities in their new environments.

```{r, fig.height = 4, fig.width=8, echo=FALSE}

act_host <- combined_data |>
  filter(
    !is.na(host_country_work_coa),
    !country %in% c("Hungary", "Bulgaria", "construct", NA)
  ) |>
  count(country, host_country_work_coa) |>
  group_by(country) |>
  mutate(percent = n / sum(n) * 100) |>
  ggplot(aes(x = country, y = percent, fill = fct_infreq(host_country_work_coa))) +
  geom_col() +
  xlab("Country") +
  ggtitle("Host Country Activities") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  ) +
  guides(fill = guide_legend(title = "Activity")) +
  scale_fill_brewer(palette = "Dark2") + theme(legend.position = "none") + ylab(NULL)

act_ukr <- combined_data|>
  filter(
    !is.na(host_country_work_coa),
    !country %in% c("Hungary", "Bulgaria", "construct", NA)
  ) |>
  filter(!is.na(demographics_resp_activity), demographics_resp_activity != "No Answer") |>
  count(country, demographics_resp_activity) |>
  group_by(country) |>
  mutate(percent = n / sum(n) * 100) |>
  ggplot(aes(x = country, y = percent, fill = demographics_resp_activity)) +
  geom_col(position = "stack") +
  xlab("Country") +
  ylab("Percent") +
  ggtitle("Employment Activity in Ukraine") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  ) +
  guides(fill = guide_legend(title = "Activity Type")) +
  scale_fill_brewer(palette = "Dark2") + theme(legend.position = "none")

shared_legend <- get_legend(act_host + theme(legend.position = "right"))

combined_plot <- plot_grid(
  act_ukr, act_host,
  labels = NULL,
  ncol = 2,
  align = "hv"
)
plot_grid(
  combined_plot, shared_legend,
  ncol = 2,  
  rel_widths = c(0.5, 0.2)
)
```
The chart illustrates the types of social protection benefits received by Ukrainian refugees across six host countries. In most countries—especially Czechia, Moldova, Romania, and Slovakia—cash benefits are the dominant form of support, reaching around 80–90% in some cases. Poland stands out as the only country where the majority (over 80%) report receiving family grants instead of cash benefits. Estonia also presents a more balanced distribution, with both cash benefits and unemployment grants being common, and a notable share receiving unemployment grants or reporting “Other Source.” Across all countries, disability and unemployment grants remain minor components. Overall, the data suggest that cash-based support remains the primary mechanism, though the specific type of aid can vary significantly by host country.

```{r, fig.height = 7, fig.width=8, echo=FALSE}
combined_df |>
  filter(!Country %in% c("Hungary", "Bulgaria", "construct", NA)) |>
    filter(Source == "benefits") |>
    mutate(Percent = round(Total_Proportion * 100, 2)) |>
    select(Component, Percent, Country) |>
    group_by(Country) |>
    arrange(desc(Percent), .by_group = TRUE) |>
    slice_head(n = 5) |>
    ungroup() |> 
  ggplot(aes(x = reorder(Component, -Percent), 
             y = Percent, fill = Component)) +
  geom_col(show.legend = TRUE) +
  facet_wrap(~Country, scales = "free_x") +
  xlab("Income Source") + ylab("Percent") +
  ggtitle("Social Protection Benefits from Host Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = guide_legend(title = "Type")) +
  scale_fill_brewer(palette = "Dark2")

```


```{r, fig.height = 7, fig.width=8, echo=FALSE}
combined_df |>
  filter(!Country %in% c("Hungary", "Bulgaria", "construct", NA)) |>
    filter(Source == "needs") |>
    mutate(Percent = round(Total_Proportion * 100, 2)) |>
    select(Component, Percent, Country) |>
    group_by(Country) |>
    arrange(desc(Percent), .by_group = TRUE) |>
    slice_head(n = 5) |>
    ungroup() |>
  ggplot(aes(x = reorder(Component, -Percent), 
             y = Percent, fill = Component)) +
  geom_col(show.legend = TRUE) +
  facet_wrap(~Country, scales = "free_x") +
  xlab("Income Source") + ylab("Percent") +
  ggtitle("Socioeconomic needs") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = guide_legend(title = "Type")) +
  scale_fill_brewer(palette = "Dark2")
```



**Safety**

According to data, 35.61% of refugees in Estonia experiences hostile
attitude, that included aggressive comments in news forums online' or
social media, verbal aggression, or physical aggression. In particular,
verbal aggression was one of the main concerns, as 14.6% of people
reported this from of hostile behavior, followed by aggressive comments
on social media and discriminatory behavior accounting for 7% and 6% of
responses respectively.

```{r}

make_prop_table(data_ee$social_cohesion_SC3_SS_EXP_HOSTILE)

combined_df |>
  filter(Source == "cohesion") |>
  mutate(Percent = round(Total_Proportion * 100, 2)) |>
  select(Component, Percent) |>
  arrange(desc(Percent))
```

**Tests**

**Random forest**

```{r, fig.width=8, fig.height=8}

set.seed(123)
important_vars <- c(
  "introduction_resp_age",
  "demographics_educ_level_grouped",
  "demographics_resp_activity",
  "income_social_protection_host_govt",
  "income_remmittances",
  "income_social_protection_ukr_govt",
  "diff_lack_of_lang",
  "diff_lack_childcare",
  "diff_lack_of_skills",
  "diff_lack_of_education_skills",
  "diff_lack_of_info",
  "diff_none",
  "diff_discrimination",
  "diff_lack_of_documentation",
  "needs_medical",
  "needs_accommodation",
  "needs_host_govt_assistance",
  "benefits_cash_benefits",
  "benefits_unemployment_grant",
  "hostile_comments_social_media",
  "hostile_verbal_aggression",
  "hostile_discriminatory_behavior",
  "hostile_none"
)

model_data <- combined_data |>
  select(employed_binary, all_of(important_vars)) |>
  drop_na()

model_data$employed_binary <- factor(model_data$employed_binary, levels = c(0,1), labels = c("No", "Yes"))

rf_model <- randomForest(employed_binary ~ ., data = model_data,
                         ntree = 500, mtry = 2, importance = TRUE)
```

```{r, fig.height=10}

imp <- importance(rf_model)
top_vars <- rownames(imp)[order(imp[, "MeanDecreaseAccuracy"], decreasing = TRUE)][1:15]

varImpPlot(rf_model)

model_data_selected <- model_data[, c("employed_binary", top_vars)]
model_data_selected$employed_binary <- as.integer(model_data_selected$employed_binary == "Yes")

logit_model <- glm(employed_binary ~ ., data = model_data_selected, family = binomial)
summary(logit_model)

pR2(logit_model)

pred_probs <- predict(logit_model, type = "response")
pred_class <- ifelse(pred_probs > 0.5, 1, 0)
table(Predicted = pred_class, Actual = model_data_selected$employed_binary)
mean(pred_class == model_data_selected$employed_binary)
roc_obj <- roc(model_data_selected$employed_binary, pred_probs)
plot(roc_obj)
auc(roc_obj)
```
**Country specific data**

```{r}

vars_pl <- c(
    "introduction_resp_age",
  "demographics_resp_activity",
  "income_social_protection_host_govt",
  "income_remmittances",
  "income_social_protection_ukr_govt",
  "diff_lack_of_lang",
  "diff_lack_childcare",
  "diff_lack_of_skills",
  "diff_lack_of_education_skills",
  "diff_none",
  "income_remmittances", 
  "diff_lack_childcare",
  "diff_not_looking_for_work",
  "diff_lack_of_age_opportunities"
)

poland_data <- combined_data |>
  filter(country == "Poland") |>
  select(employed_binary, all_of(vars_pl)) |>
  drop_na()
poland_data$employed_binary <- as.integer(poland_data$employed_binary == 1)

logit_model_pl <- glm(employed_binary ~ ., data = poland_data, family = binomial)
summary(logit_model_pl)

pR2(logit_model_pl)

pred_probs_pl <- predict(logit_model_pl, type = "response")
pred_class_pl <- ifelse(pred_probs_pl > 0.5, 1, 0)

# Confusion matrix and accuracy
table(Predicted = pred_class_pl, Actual = poland_data$employed_binary)
mean(pred_class_pl == poland_data$employed_binary)
roc_obj_pl <- roc(poland_data$employed_binary, pred_probs_pl)
plot(roc_obj_pl)
auc(roc_obj_pl)

```

```{r}

vars_ee <- c(
  "demographics_resp_activity",
  "diff_lack_of_lang",
  "diff_lack_of_documentation",
  "benefits_cash_benefits",
  "benefits_unemployment_grant",
  "hostile_comments_social_media",
  "hostile_verbal_aggression",
  "hostile_discriminatory_behavior",
  "diff_loss_benefits",
  "needs_employment"
)

estonia_data <- combined_data |>
  filter(country == "Estonia") |>
  select(employed_binary, all_of(vars_ee)) |>
  drop_na()
estonia_data$employed_binary <- as.integer(estonia_data$employed_binary == 1)

logit_model_ee <- glm(employed_binary ~ ., data = estonia_data, family = binomial)
summary(logit_model_ee)

pR2(logit_model_ee)

pred_probs_ee <- predict(logit_model_ee, type = "response")
pred_class_ee <- ifelse(pred_probs_ee > 0.5, 1, 0)

# Confusion matrix and accuracy
table(Predicted = pred_class_ee, Actual = estonia_data$employed_binary)
mean(pred_class_ee == estonia_data$employed_binary)
roc_obj_ee <- roc(estonia_data$employed_binary, pred_probs_ee)
plot(roc_obj_ee)
auc(roc_obj_ee)

```

```{r}

vars_cz <- c(
  "demographics_resp_activity",
  "income_social_protection_host_govt",
  "diff_lack_of_lang",
  "diff_none",
  "hostile_verbal_aggression",
  "diff_lack_of_decent_employment",
  "diff_lack_childcare",
  "diff_not_looking_for_work"
)

czechia_data <- combined_data |>
  filter(country == "Czechia") |>
  select(employed_binary, all_of(vars_cz)) |>
  drop_na()
czechia_data$employed_binary <- as.integer(czechia_data$employed_binary == 1)

logit_model_cz <- glm(employed_binary ~ ., data = czechia_data, family = binomial)
summary(logit_model_cz)

pR2(logit_model_cz)

pred_probs_cz <- predict(logit_model_cz, type = "response")
pred_class_cz <- ifelse(pred_probs_cz > 0.5, 1, 0)

table(Predicted = pred_class_cz, Actual = czechia_data$employed_binary)
mean(pred_class_cz == czechia_data$employed_binary)

roc_obj_cz <- roc(czechia_data$employed_binary, pred_probs_cz)
plot(roc_obj_cz)
auc(roc_obj_cz)

```

```{r}


vars_sk <- c(
  "introduction_resp_age",
  "demographics_educ_level_grouped",
  "demographics_resp_activity",
  "income_social_protection_host_govt",
  "diff_lack_of_skills",
  "diff_not_looking_for_work",
  "needs_host_govt_assistance",
  "needs_humanitarian_assistance",
  "diff_none",
  "needs_host_govt_assistance",
  "benefits_cash_benefits"
)

slovakia_data <- combined_data |>
  filter(country == "Slovakia") |>
  select(employed_binary, all_of(vars_sk)) |>
  drop_na()
slovakia_data$employed_binary <- as.integer(slovakia_data$employed_binary == 1)

logit_model_sk <- glm(employed_binary ~ ., data = slovakia_data, family = binomial)
summary(logit_model_sk)

pR2(logit_model_sk)

pred_probs_sk <- predict(logit_model_sk, type = "response")
pred_class_sk <- ifelse(pred_probs_sk > 0.5, 1, 0)

table(Predicted = pred_class_sk, Actual = slovakia_data$employed_binary)
mean(pred_class_sk == slovakia_data$employed_binary)

roc_obj_sk <- roc(slovakia_data$employed_binary, pred_probs_sk)
plot(roc_obj_sk)
auc(roc_obj_sk)

```

```{r}


vars_ml <- c(
  "demographics_resp_activity",
  "income_social_protection_host_govt",
  "diff_lack_of_lang",
  "diff_lack_of_skills",
  "diff_lack_of_info",
  "diff_none",
  "needs_medical",
  "diff_need_to_take_care_of_others",
  "needs_humanitarian_assistance"
)

moldova_data <- combined_data |>
  filter(country == "Moldova") |>
  select(employed_binary, all_of(vars_ml)) |>
  drop_na()
moldova_data$employed_binary <- as.integer(moldova_data$employed_binary == 1)

logit_model_md <- glm(employed_binary ~ ., data = moldova_data, family = binomial)
summary(logit_model_md)

pR2(logit_model_md)

pred_probs_md <- predict(logit_model_md, type = "response")
pred_class_md <- ifelse(pred_probs_md > 0.5, 1, 0)

table(Predicted = pred_class_md, Actual = moldova_data$employed_binary)
mean(pred_class_md == moldova_data$employed_binary)

roc_obj_md <- roc(moldova_data$employed_binary, pred_probs_md)
plot(roc_obj_md)
auc(roc_obj_md)
```

```{r}


vars_ro <- c(
  "introduction_resp_age",
  "demographics_resp_activity",
  "diff_lack_of_lang",
  "diff_lack_of_decent_employment",
  "diff_lack_childcare",
  "diff_none",
  "needs_host_govt_assistance","income_other_sources",
  "diff_lack_of_skills",
  "diff_need_to_take_care_of_others",
  "diff_not_looking_for_work"
)

romania_data <- combined_data |>
  filter(country == "Romania") |>
  select(employed_binary, all_of(vars_ro)) |>
  drop_na()
romania_data$employed_binary <- as.integer(romania_data$employed_binary == 1)

logit_model_ro <- glm(employed_binary ~ ., data = romania_data, family = binomial)
summary(logit_model_ro)

pR2(logit_model_ro)

pred_probs_ro <- predict(logit_model_ro, type = "response")
pred_class_ro <- ifelse(pred_probs_ro > 0.5, 1, 0)

table(Predicted = pred_class_ro, Actual = romania_data$employed_binary)
mean(pred_class_ro == romania_data$employed_binary)

roc_obj_ro <- roc(romania_data$employed_binary, pred_probs_ro)
plot(roc_obj_ro)
auc(roc_obj_ro)
```

*Discussion*

Based on the output of logistic regression, reporting no difficulties increases the odds of employment by 768%, indicating that individuals who report no barriers are far more likely to be employed. Lack of language skills surprisingly increases employment odds by 223%, likely reflecting that those actively seeking work are both employed and aware of language as a barrier - suggesting this may be a sign for effort or exposure. Lack of skills increases odds by 56%, and lack of education skills by 70%, again potentially reflecting motivated individuals who are aware of their deficits but engaged in the labor market. Receiving unemployment benefits is associated with a 78% decrease in odds of employment, and cash benefits reduce odds by 50%, suggesting potential disincentives or that recipients are in more vulnerable situations. Being unemployed in Ukraine reduces the odds of employment in the European country of residence by 58% compared to the reference activity group, showing the persistence of non-employment status.
Individuals needing host government assistance, medical, or accommodation support have reduced employment odds by 56%, 39%, and 16% respectively, highlighting structural barriers to integration. Older adults (60+) are 65% less likely to be employed than those under 35. People doing housekeeping, studying, or retired have significantly lower employment odds (around 55–85% lower) compared to the baseline group.

*Conclusion*

*Citations*

- SocioFactor, IOM, Impact-REACH, SHC, Sociofactor, TARKI Social Research Institute, Ipsos, UNHCR (2023). Poland, Slovak Republic, Hungary, Czech Republic, Moldova, Romania, Bulgaria: Multi-Sector Needs Assessment (MSNA) - 2023. Accessed from: https://microdata.unhcr.org

- UNHCR (2023). Estonia: Multi-Sector Needs Assessment (MSNA) - 2023. Accessed from: https://microdata.unhcr.org

- Preut, Holger. “고용을위한길:독일노동시장의우크라이나난민들을위한통합적도전과전망.” Zeitschrift Der Koreanisch-Deutschen Gesellschaft Für Sozialwissenschaften, vol. 33, no. 3, Koreanisch-Deutsche Gesellschaft fuer Sozlaiwssenschaften, Sept. 2023, pp. 3–33. Crossref, doi:10.19032/zkdgs.2023.09.33.3.3.

- Londar, S., et al. “Challenges for Ukrainian Refugees in the EU Labour Market: The Case of Poland and Germany.” Educational Analytics of Ukraine, no. 5, State Scientific Institution - Institute of Educational Analytics, 2024, pp. 5–16. Crossref, doi:10.32987/2617-8532-2024-5-5-16.
